Hurricane Evacuation Strategies for the State of Mississippi

### 2015-11-16 manuscript.tex
Wamp wamp
Added some things to contributions
Brought back cropped

\section{Introduction}
\label{sec:introduction}
  The Mississippi Emergency Management Administration (MSEMA) has hired our team to design hurricane evacuation routes for at-risk counties in the coastal regions of southern Mississippi. Given that this coastal area is a strong commercial center, the government authorities want us to minimize the number of unnecessary evacuations. Under this requirement, it is more advantageous to withhold ordering an evacuation for as long as safely possible. The reasons for this are two-fold. First, this ensures that local businesses can remain open for as long as possible. Second, this ensures that the forecasts at the time of the evacuation decision are as accurate as possible. Therefore, our models seek to recommend an optimal evacuation plan as well as the latest possible time at which it is safe to enact this plan.

\section{Network Representation of Mississippi}
\label{sec:representation}
  \par
    The MSEMA specified that our plan should be enacted on a per-county basis. As you can see in the following figure, these counties are approximately equally sized. As a result of this, we assume that each county can be modeled as a single node in a bidirectional graph. Then an edge from node $i$ to node $j$ represents an available evacuation route from county $i$ to county $j$. In all of our models we assume that highways are the only means of evacuation. Therefore, we constructed a network of Mississippi counties by connecting each county's node to the nodes of all bordering counties directly reachable by highways:
  \begin{figure}[H]
    \center
    \begin{subfigure}[b]{0.5\textwidth}
      \center
      \includegraphics[width=.9\linewidth]{figures/county_map.jpg}
      \caption*{A Map of Counties in the State of Mississippi \cite{county_map}}
    \end{subfigure}~
    \begin{subfigure}[b]{0.5\textwidth}
      \center
      \includegraphics[width=\linewidth]{figures/full_undirected-crop.pdf}
      \caption*{Undirected Graph Representing Mississippi Counties}
    \end{subfigure}
    \caption*{}
  \end{figure}
  \par
     Because Mississippi has varying highway structure, it would be inappropriate to assume uniform travel rates between county pairs. In order to capture this heterogeneity in the network model, we decided to assign weights to every edge. Then a single edge weight represents a carrying capacity for the amount of travel possible between two counties. For a real-world mapping, we assume that this would correspond to the total number of highway lanes spanning two counties. Therefore, we were able to assign edge weights across the entire graph by using the following highway map:
  \begin{figure}[H]
    \centering
    \includegraphics[width=.5\textwidth]{ms_highways.pdf}
    \caption{Map of Major Highways in the State of Mississippi \cite{ms_highways}}
    \label{fig:highway_map}
  \end{figure}
  \par
    In this figure, the thick black lines represent four lane highways while the red lines represent two lane highways. We assume that despite the ensuing panic, evacuees still travel along the right side of the highway, even if this means lanes going the other way (towards the hurricane) are underutilized. Therefore, for every thick black line spanning two counties, the appropriate edge in the graph gains two additional lanes, whereas red lines add one lane. Summing the lane contributions of all highways for a pair of counties therefore gives us an single edge weight in the graph.
  \newline
  \par
    We assume that there is a fixed spacing between every car on every highway. Then all cars travel at the same speed, and the density of cars is uniform along the entire length of every highway. This determines a fixed flow rate per edge capacity, and allows us to compute the number of cars traveling to another county in a given time interval. Because of the inevitable traffic congestion during an evacuation, we set the fixed speed to a relatively slow $10$ km/hr. In a traffic jam situation like this, we expect a tight traffic density of one car every 10 meters. Therefore:
    \[
      \frac{10\ \text{km/hr}}{10\ \text{m/car}} = 1000\ \text{cars/hr}
    \]
  \par
    We assume that evacuees travel with friends and family in their cars. Because the average size of an American family is \texttildelow 3 persons \cite{famsize}, we conclude that 3000 people per hour can use a single lane to another county. Maximum population flow rates between each county are therefore three thousand times the corresponding edge weight.

@@

  \subsection{Parameter Values and Justification}
    Using our earlier values for the maximum flow rate of an edge, we can now use them to derive transition probabilities for our Markov model. Because the maximum capacity of a road is 3000 times it's weight in the graph, this can be used to give the total population emanating from a county in the first hour of the simulation. If we find the ratios of these values along each edge to the initial population in the county, this gives us a good approximation of the percentage of a county that flees in any given time step. Again we assume these proportions are constant for each edge and ensure conservation of probability through the whole graph. We ensured this conservation of probability by ensuring a constant population throughout the state at every time step.\\
    \\
    In order to get the threshold for each county based on the hurricane characteristics, we created a function that would give us the proportion of the original population that did not need to be evacuated. For instance, during a category 1 hurricane, we would find it acceptable to only evacuate 50\% of the original population. Similarly, the proximity of the county in question should factor into that threshold.
    The formula we ended up using is the following:
    \begin{align*}
        threshold(c, h) = 0.1 \cdot 2^{||c-h||_2 / h.radius} \cdot \frac{1}{h.category}
    \end{align*}
    In addition to this, if a county is further away than 3 times the radius of the hurricane, that county does not need to be evacuated, and we tag it as a refugee state (it essentially has infinite threshold). Furthermore, category 1 hurricanes have a threshold minimum of 50\% (we do not require more than half of the population of any county to evacuate).
    The heatmaps for those thresholds can be observed in Figure~\ref{fig:thresholds}. These were generated from a category 1 hurricane (a) that then grows in size (b) and finally becomes category 5 (c).
    \begin{figure}[H]
      \center
      \begin{subfigure}[b]{0.3\textwidth}
        \center
        \includegraphics[width=\textwidth]{figures/right_73_1-crop.pdf}
        \caption{Radius: 73km - Category 1}
        \label{fig:pop_county60}
      \end{subfigure}~
      \begin{subfigure}[b]{0.3\textwidth}
        \center
        \includegraphics[width=\textwidth]{figures/right_120_1-crop.pdf}
        \caption{Radius: 120km - Category 1}
        \label{fig:pop_county39}
      \end{subfigure}~
      \begin{subfigure}[b]{0.3\textwidth}
        \center
        \includegraphics[width=\textwidth]{figures/right_120_5-crop.pdf}
        \caption{Radius: 120km - Category 5}
        \label{fig:pop_county60}
      \end{subfigure}
      \caption{Thresholds of evacuation for counties of Mississippi}
      \label{fig:thresholds}
    \end{figure}

@@

\subsection*{Antoine}
Implemented the flow and Markov models. Created almost all of the graph and network visualizations. Worked on the two first models' analyses for the manuscript.

>>>>

\section{Introduction}
\label{sec:introduction}
  The Mississippi Emergency Management Administration (MSEMA) has hired our team to design hurricane evacuation routes for at-risk counties in the coastal regions of southern Mississippi. Given that this coastal area is a strong commercial center, the government authorities want us to minimize the number of unnecessary evacuations. Under this requirement, it is more advantageous to withhold ordering an evacuation for as long as is safely possible. The reasons for this are two-fold. First, this ensures that local businesses can remain open for as long as possible. Second, this ensures that the forecasts at the time of the evacuation decision are as accurate as possible. Therefore, our models seek to recommend an optimal evacuation plan as well as the latest possible time at which it is safe to enact this plan.

\section{Network Representation of Mississippi}
\label{sec:representation}
  \par
    The MSEMA specified that our plan should be enacted on a per-county basis. As you can see in the following figure, these counties are approximately equally sized. As a result of this, we assume that each county can be modeled as a single node in a bidirectional graph. Then an edge from node $i$ to node $j$ represents an available evacuation route from county $i$ to county $j$. In all of our models we assume that highways are the only means of evacuation. Therefore, we constructed a network of Mississippi counties by connecting each county's node to the nodes of all bordering counties directly reachable by highways:
  \begin{figure}[H]
    \center
    \begin{subfigure}[b]{0.5\textwidth}
      \center
      \includegraphics[width=.9\linewidth]{figures/county_map.jpg}
      \caption*{A Map of Counties in the State of Mississippi \cite{county_map}}
    \end{subfigure}~
    \begin{subfigure}[b]{0.5\textwidth}
      \center
      \includegraphics[width=\linewidth]{figures/full_undirected.pdf}
      \caption*{Undirected Graph Representing Mississippi Counties}
    \end{subfigure}
    \caption*{}
  \end{figure}
  \par
     Because Mississippi has varying highway structure, it would be inappropriate to assume uniform travel rates between county pairs. In order to capture this heterogeneity in the network model, we decided to assign weights to every edge. Then a single edge weight represents a carrying capacity for the amount of travel possible between two counties. For a real-world mapping, we assume that this would correspond to the total number of highway lanes spanning two counties. Therefore, we were able to assign edge weights across the entire graph by using the following highway map:
  \begin{figure}[H]
    \centering
    \includegraphics[width=.5\textwidth]{ms_highways.pdf}
    \caption{Map of Major Highways in the State of Mississippi \cite{ms_highways}}
    \label{fig:highway_map}
  \end{figure}
  \par
    In this figure, the thick black lines represent four lane highways while the red lines represent two lane highways. We assume that despite the ensuing panic, evacuees still travel along the right side of the highway, even if this means lanes going the other way (towards the hurricane) are underutilized. Therefore, for every thick black line spanning two counties, the appropriate edge in the graph gains two additional lanes, whereas red lines add one lane. Summing the lane contributions of all highways for a pair of counties therefore gives us an single edge weight in the graph.
  \newline
  \par
    We assume that there is a fixed spacing between every car on every highway. Then all cars travel at the same speed, and the density of cars is uniform along the entire length of every highway. This determines a fixed flow rate per edge capacity, and allows us to compute the number of cars traveling to another county in a given time interval. Because of the inevitable traffic congestion during an evacuation, we set the fixed speed to a relatively slow $10$ km/hr. In a traffic jam situation like this, we expect a tight traffic density of one car every 10 meters. Therefore:
    \[
      \frac{10\ \text{km/hr}}{10\ \text{m/car}} = 1000\ \text{cars/hr}
    \]
  \par
    We assume that evacuees travel with friends and family in their cars. Because the average size of an American family is \texttildelow 3 persons \cite{famsize}, we conclude that 3000 people per hour can use a single lane to another county. Maximum population flow rates between each county are therefore three thousand times the corresponding edge weight.

@@

  \subsection{Parameter Values and Justification}
    Using our earlier values for the maximum flow rate of an edge, we can now use them to derive transition probabilities for our Markov model. Because the maximum capacity of a road is 3000 times it's weight in the graph, this can be used to give the total population emanating from a county in the first hour of the simulation. If we find the ratios of these values along each edge to the initial population in the county, this gives us a good approximation of the percentage of a county that flees in any given time step. Again we assume these proportions are constant for each edge and ensure conservation of probability through the whole graph. We ensured this conservation of probability by ensuring a constant population throughout the state at every time step.\\
    \\
    In order to get the threshold for each county based on the hurricane characteristics, we created a function that would give us the proportion of the original population that did not need to be evacuated. For instance, during a category 1 hurricane, we would find it acceptable to only evacuate 50\% of the original population. Similarly, the proximity of the county in question should factor into that threshold.
    The formula we ended up using is the following:
    \begin{align*}
        threshold(c, h) = 0.1 \cdot 2^{||c-h||_2 / h.radius} \cdot \frac{1}{h.category}
    \end{align*}
    In addition to this, if a county is further away than 3 times the radius of the hurricane, that county does not need to be evacuated, and we tag it as a refugee state (it essentially has infinite threshold). Furthermore, category 1 hurricanes have a threshold minimum of 50\% (we do not require more than half of the population of any county to evacuate).
    The heatmaps for those thresholds can be observed in Figure~\ref{fig:thresholds}. These were generated from a category 1 hurricane (a) that then grows in size (b) and finally becomes category 5 (c).
    \begin{figure}[H]
      \center
      \begin{subfigure}[b]{0.3\textwidth}
        \center
        \includegraphics[width=\textwidth]{figures/right_73_1.pdf}
        \caption{Radius: 73km - Category 1}
        \label{fig:pop_county60}
      \end{subfigure}~
      \begin{subfigure}[b]{0.3\textwidth}
        \center
        \includegraphics[width=\textwidth]{figures/right_120_1.pdf}
        \caption{Radius: 120km - Category 1}
        \label{fig:pop_county39}
      \end{subfigure}~
      \begin{subfigure}[b]{0.3\textwidth}
        \center
        \includegraphics[width=\textwidth]{figures/right_120_5.pdf}
        \caption{Radius: 120km - Category 5}
        \label{fig:pop_county60}
      \end{subfigure}
      \caption{Thresholds of evacuation for counties of Mississippi}
      \label{fig:thresholds}
    \end{figure}

@@

\subsection*{Antoine}
Implemented the flow and Markov models. Created almost all of the graph and network visualizations. Worked on the two first models' analyses for the manuscript. Wrote the Future Work section and the Individual Contributions.
